Studies of quantum dots

### 2017-01-14 Manuscript/paper.tex
Update spEnergiesQD/Manuscript/paper.tex

We model the quantum dot as a collection of $N$ electrons trapped in an
external harmonic-oscillator potential.\footnote{In this paper, we use atomic
  units where $\hbar$, $m$, $e$, and $4 \pi \epsilon$ are set to 1.  Here,
  $\epsilon$ denotes the permittivity and $m$ denotes the mass, which could be
  an effective mass rather than the true electron mass.  In effect, the unit
  of energy in this article is
  $E_{\text h} (m / m_{\text e}) / (\epsilon / \epsilon_0)^2$ where
  $E_{\text h} \equiv m_{\text e} (\alpha c)^2$, the hartree unit of energy.}
Each electron possesses kinetic energy $t(\bm p) \equiv \bm p^2 / 2$ and
experiences an external potential $u(\bm r) \equiv \omega^2 \bm r^2 / 2$,
where $\omega$ is the angular frequency of the oscillator, $\bm r$ is the
2-dimensional position vector of an electron, and $\bm p$ is its linear
momentum.  Each pair of electrons experience a Coulomb repulsion
$v(\bm r, \bm r') \equiv |\bm r - \bm r'|^{-1}$ where $\bm r$ and $\bm r'$ are
their positions.  The full many-body problem is described by the
nonrelativistic Hamiltonian
\begin{align*}
  \hat H \equiv \hat H^{\text o} + \hat V = \sum_j \bigl(
    t(\hat{\bm p}_j) + u(\hat{\bm r}_j)
  \bigr) + \sum_{j_1 < j_2} v(\hat{\bm r}_{j_1}, \hat{\bm r}_{j_2})
\end{align*}
where $\hat{\bm p}_j$ and $\hat{\bm r}_j$ respectively denote the momentum and
position operators of a particle $j$.  The first summation is performed over
every particle, while the the second summation is performed over every
unordered pair of distinct particles (the ordering of the particle labels $j$ is arbitrary).

The full problem is not a trivial one.  We shall first consider a much simpler problem: the noninteracting part of the Hamiltonian, $\hat H^{\text o}$, which does not contain the Coulomb interaction $v$:
\begin{align} \label{eq:onebodyhamiltonian}
  \hat H^{\text o} \equiv \sum_p \hat h^{\text o}_p
  \equiv \sum_{p} \bigl(
  t(\hat{\bm p}_p) + u(\hat{\bm r}_p)
  \bigr)
\end{align}
In this case, the many-body Hamiltonian reduces to $N$ independent
single-particle Hamiltonians, each of the form:
\begin{align*}
  \hat h^{\text o} \equiv
  \frac{\hat{\bm p}^2 + \omega^2 \hat{\bm r}^2}{2}
\end{align*}
This is a 2-dimensional harmonic oscillator problem for a single particle,
whose solutions are well known.  There are various ways to construct the set
of eigenfunctions: we shall use the Fock-Darwin states that conserve angular
momentum $\hat{L}_{\text{z}}$ about the $z$-axis (perpendicular to the 2D
plane).  The eigenfunctions have the following analytic form in polar
coordinates:\citep[Appx.\ A]{lohne2010coupled}
\begin{gather*}
  \phi^{\text{FD}}_{n, m_\ell}(\rho, \varphi)
  \equiv \sqrt\omega
  \phi^{\text{FDr}}_{n, |m_\ell|}(\sqrt \omega \rho)
  \phi^{\text{Lz}}_{m_\ell}(\varphi)
\end{gather*}
where $\rho$ denotes the radial coordinate, $\varphi$ denotes the angular coordinate,
\begin{align*}
  \phi^{\text{FDr}}_{n, m}(r) &\equiv
  \sqrt{\frac{2 \cdot n!}{(n + m)!}}
  {\mathrm e}^{-r^2 / 2} r^m L_n^{(m)}(r^2) \\
  \phi^{\text{Lz}}_{m_\ell}(\varphi) &\equiv
  \frac{1}{\sqrt{2 \pi}} {\mathrm e}^{i m_\ell \varphi}
\end{align*}
and $L_n^{(\mu)}$ denotes the generalized Laguerre polynomial of degree $n$
with parameter $\mu$.  The states are labeled by two quantum numbers: the
principal quantum number $n \in \{0, 1, \ldots\}$ and the angular momentum
projection $m_\ell \in \mathbb Z$.

@@

\begin{figure}
\includegraphics[width=.48\textwidth]{fig-shell-structure}
\caption{Structure of the lowest 12 single-particle states in a noninteracting
  harmonic oscillator of frequency $\omega = 1$.  Each box represents a
  single-particle state, and the arrows denote the spin of the states.}
\label{fig:shell-structure}
\end{figure}
The eigenvalues of the single-particle Hamiltonian $\hat h^{\text o}$ are
given by
\begin{align*}
  \varepsilon_{n, m_\ell, m_{\text s}} \equiv (2 n + |m_\ell| + 1) \omega
\end{align*}
which satisfies
$\hat{h}^{\text o} \ket{\phi_{n, m_\ell, m_{\text s}}} = \varepsilon_{n,
  m_\ell, m_{\text s}} \ket{\phi_{n, m_\ell, m_{\text s}}}$.  Note that the
states are degenerate with respect to the spin projection $m_{\text s}$ since
our Hamiltonian $\hat h^{\text o}$ does not distinguish between them.  From
this equation, we can deduce the shell structure as shown in Figure
\ref{fig:shell-structure}.  It is notable that there are degeneracies beyond
that of the spin pairs: each shell contains a set of states that satisfy the
equation $2 n + |m_\ell| = k$ for some nonnegative integer $k$, which we
conveniently call the \textit{shell index}.  The shells are equidistant from
each other, with a spacing equal to the frequency $\omega$.

The many-body Hamiltonian $\hat H^{\text o}$ is the sum of the single-particle
Hamiltonians $\hat h^{\text o}_p$ over all particles $p$, where $p$ may now be
treated as a tuple of all three single-particle quantum numbers:
$(n, m_\ell, m_{\text s})$.  From the states $\phi$ in
\eqref{eq:singleparticlestate} we can construct the an eigenstate of the
many-body Hamiltonian $\hat H^{\text o}$ in the form of a Slater determinant
$\varnothing_{p_1, \ldots, p_N}$:
\begin{align*}
  \varnothing_{p_1, \ldots, p_N}(\bm r_1, \ldots, \bm r_N) \equiv
  \frac{1}{\sqrt{N}} \left|
  \begin{matrix}
    \phi_{p_1}(\bm r_1) & \cdots & \phi_{p_N}(\bm r_1) \\
    \vdots & \ddots & \vdots \\
    \phi_{p_1}(\bm r_N) & \cdots & \phi_{p_N}(\bm r_N) \\
  \end{matrix}
  \right|
\end{align*}
This automatically satisfies the Pauli exclusion principle, ensuring that the
Fermi statistics are satisfied.  The energy of these states is given by the
sum of the single-particle energies:
\begin{align*}
  E^{\text o}_{p_1, \ldots, p_N} =
  \varepsilon_{p_1} + \cdots + \varepsilon_{p_N},
\end{align*}
which satisfies
$\hat{H}^{\text o} \ket{\varnothing_{p_1, \ldots, p_N}} = E^{\text o}_{p_1, \ldots,
  p_N} \ket{\varnothing_{p_1, \ldots, p_N}}$.  For ground states, we may have
degeneracies due to the shell structure of the single-particle basis.
However, if $N = k_{\text F} (k_{\text F} + 1)$ for some nonnegative integer
$k_{\text F}$, there are just enough particles to form a closed-shell system.
In this case, there is a unique, well-isolated ground state.  The values of
$N$ for which this occurs are typically referred to as \textit{magic numbers}
and we call $k_{\text F}$ the \textit{number of filled shells} (or ``Fermi
level'').  This completes the solution of the noninteracting problem.

@@

Operators can be rewritten in this formalism as well.  The one-body
Hamiltonian $H^{\text o}$ defined by \eqref{eq:onebodyhamiltonian} can be
rewritten in second quantization as
\begin{align} \label{eq:second_quantized_one_body_hamiltonian}
  \hat H^{\text o} = \sum_{p, q} H_{p, q}^{\text o} \hat a_p^\dagger \hat a_q^{}
\end{align}
where the \textit{matrix element}
$H_{p, q}^{\text o} \equiv \bra{\varnothing_p} \hat H^{\text o} \ket{\varnothing_q}$.  This
can likewise be extended to $\hat V$,
\begin{align} \label{eq:second_quantized_two_body_hamiltonian}
  \hat V = \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s}^{}
  \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}
\end{align}
where $U_{p, q} \equiv \langle \varnothing_p | \hat U | \varnothing_q \rangle$ and
$V_{p, q, r, s} \equiv \langle \varnothing_{p, q} | \hat V | \varnothing_{r, s} \rangle$.
The complete Hamiltonian is given by their sum:
\begin{align}
  \hat{H} = \sum_{p, q} H^{\text o}_{p, q} \hat{a}_p^\dagger \hat{a}_q + \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_r \hat{a}_s  \label{eq:vacuumhamiltonian}
\end{align}
Note that due to the antisymmetry present in the Slater determinants, we have
the property that exchanging either $p$ with $q$ or $r$ with $s$ causes
$V_{p, q, r, s}$ to reverse in sign, leading to a $4$-fold symmetry that is
canceled out by the factor of $1/4$.  For this reason, the two-body matrix
element $V_{p, q, r, s}$ is commonly referred to as the \emph{antisymmetrized}
matrix element.  This is not to be confused with the non-antisymmetric
interaction integral defined by:
\begin{align*}
  \tilde V_{p, q, r, s} \equiv \int \phi_p(\bm r) \phi_q(\bm r') v(\bm r, \bm r') \phi_r(\bm r) \phi_s(\bm r') \D \bm r \D \bm r'
\end{align*}
The antisymmetrized matrix element is computed from the interaction integral
by $V_{p, q, r, s} = \tilde V_{p, q, r, s} - \tilde V_{p, q, s, r}$.  In our
work, we compute the interaction numerically via the OpenFCI software, which
numerical integrates the Coulomb interaction in the center-of-mass frame
[[cite Kvaal's paper; also, is the description correct?]].

@@

Since the coefficient matrix uniquely determines the unknown states, the variational problem has been reduced to that of solving of the coefficients $\bm C$ that minimize the energy $E_{\Phi}$, which is given by
\begin{align}
  E_{\Phi} &= \sum_{j} n_j H^{\text{o} \prime}_{j, j} + \sum_{j, k} n_j n_k V'_{j, k, j, k} \label{eq:hfenergy}
\end{align}
where the transformed operators are defined as
\begin{align}
  H^{\text{o} \prime}_{p, q} &\equiv \sum_{r, s} C_{r, p}^* H_{r, s}^{\text o} C_{s, q}^{} \label{eq:hftransform1} \\
  V'_{p, q, r, s} &\equiv \sum_{t, u, v, w} C_{t, p}^* C_{u, q}^* V_{t, u, v, w}^{} C_{v, r}^{} C_{w, s}^{} \label{eq:hftransform2}
\end{align}
and the occupancy number $n_p$ is defined as
\begin{align}
  n_p \equiv \begin{cases}
    1 & \text{if $p$ is an occupied state in $\Phi$} \\
    0 & \text{if $p$ is an unoccupied state in $\Phi$}
  \end{cases}
\end{align}
and further define $\bar n_p \equiv 1 - n_p$ as the complement.

With the method of Lagrange multipliers, the minimization problem can be reduced to the solving of a nonlinear equation -- the \textit{Hartree-Fock equation}:
\begin{align} \label{eq:hartreefock}
  \bm F \bm C = \bm C \bm \epsilon
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p, q} \equiv H_{p, q}^{\text o} + \sum_{r, s, j} n_j C_{r, j}^* V_{p, r, q, s}^{} C_{s, j}^{}
\end{align}
and $\bm \epsilon$ is a vector of Lagrange multipliers with each element $\epsilon_q$ associated with a particular state $\phi'_q$.

@@

To make use of the HF solution as the reference state for future calculations, we transform the operators via \eqref{eq:hftransform1} and \eqref{eq:hftransform2}.  Therefore, the post-HF methods described in the next few sections generally use the transformed operators $\hat H^{\text{o} \prime}$ and $\hat V'$ as inputs.  However, for generality, we omit the prime symbols as the methods do not necessitate a state optimized by HF.

@@

We may then define \emph{normal ordering with respect to the reference state} $\Phi$, a notational device that rearranges a string of operators such that every quasiparticle creation operator appears before a quasiparticle annihilation operator by repeatedly exchanging neighboring operators, with each exchange incurring a change in sign.  A simple example would be the normal ordering of a pair of field operators such as:
\begin{align*}
  \normord{\hat{b}_p \hat{b}_q^\dagger} =
  -\hat{b}_q^\dagger \hat{b}_p
\end{align*}
Here, we denote normal ordering of an operator string with respect to $\Phi$ by a pair of colons.  The expectation value of a normal ordered string of operators with respect to $\Phi$ is guaranteed to vanish:
\begin{align*}
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q \ket{\Phi} = 0 &
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q^\dagger \hat{b}_s \hat{b}_r \ket{\Phi} = 0
\end{align*}
The normal ordering of the quasiparticle field operators are straightforward to compute, but what is more interesting is the normal ordering of the physical field operators $\hat a$ and $\hat a^\dagger$ with respect to $\Phi$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q} &=
  \hat{a}_p^\dagger \hat{a}_q - n_p \delta_{p, q}
\end{align*}
The normal ordering notation thus provides a concise way of expressing the right-hand side.  In this new notation, we can rewrite the Hamiltonian $\hat H = \hat H^{\text o} + \hat V$ in its \emph{normal-ordered representation}:
\begin{align}
  \hat{H} = H^{[0]} + \sum_{p, q} H^{[1]}_{p, q} \normord{\hat{a}_p^\dagger \hat{a}_q} + \frac{1}{4} \sum_{p, q, r, s} H^{[2]}_{p, q, r, s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_r \hat{a}_s}
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  H^{[0]} &\equiv \sum_{i} n_i H^{\text o}_{i, i} + \frac{1}{2} \sum_{i, j} n_i n_j V_{i, j, i, j} \\
  H^{[1]}_{p, q} &\equiv H^{\text o}_{p, q} + \sum_{i} n_i V_{p, i, q, i} \\
  H^{[2]}_{p, q, r, s} &\equiv V_{p, q, r, s} \\
\end{align*}
If $\Phi$ is a state optimized by the HF method, then $H^{[0]}$ is simply the HF energy $E_\Phi$ and $H^{[1]}_{p, q}$ is the Fock matrix $F_{p, q}$.  The normal-ordered representation of $\hat H$ in \eqref{eq:normordhamiltonian} is completely equivalent to that of $\hat H$ in \eqref{eq:vacuumhamiltonian}.  The only difference is that it rearranges the matrix elements so as to emphasize the reference state $\Phi$ rather than vacuum state $\ket{\varnothing}$.  This makes a critical difference when the expressions containing 3-body or higher terms are \emph{truncated} to only 2-body.  If the reference state $\Phi$ is good approximation of the true solution $\Psi$, then the importance of the higher-body parts of the Hamiltonian are greatly reduced.

@@

In QDPT, the solutions of the approximate Hamiltonian $\hat H^{\text o}$ are used as the basis of the model space.  One then assumes the existence of an operator $\hat \Omega$, known as the \textit{wave operator}, that maps some set of states $\tilde \Psi^{\text o}_u$ within the model space to the exact ground state $\Psi_u$:
\begin{align} \label{eq:omega-condition1}
  \Psi_u = \hat \Omega \tilde \Psi^{\text o}_u
\end{align}
The states $\tilde \Psi^{\text o}_u$ consist of some mixture of the eigenstates
$\Psi^{\text o}_{u'}$ of the approximate Hamiltonian $\hat H^{\text o}$.

There is some freedom in the choice of the wave operator $\hat \Omega$.  We
assume it has the following form:
\begin{align} \label{eq:omega-condition2}
  \hat \Omega = \hat P + \hat Q \hat \Omega \hat P
\end{align}
where $\hat P$ projects any state into the model space and $\hat Q$ is the
complement of $\hat P$.  This entails that the exact states $\Psi_u$ is no
longer normalized but instead satisfies the typical intermediate
normalization: $\langle \Psi_u | \tilde \Psi^{\text o}_u \rangle = 1$.

>>>>

We shall model the circular quantum dot system as a collection of $N$ nonrelativistic electrons of mass $m$ in 2-dimensional space, trapped by an external harmonic-oscillator potential of the form $m \omega^2 r^2 / 2$, where $\omega$ is the angular frequency of the oscillator trap and $r$ is the distance from the center of the trap.  The electrons interact with each other through the standard Coulomb interaction $e^2 / (4 \pi \epsilon R)$, where $e$ is the charge of each electron, $\epsilon$ is the permittivity of the medium, and $R$ is the distance between the two interacting electrons.

Even though the model system contains 5 parameters $(N, m, \omega, e, \epsilon)$, many of them are redundant.  With an appropriate choice of units, the number of parameters in the system can be reduced to just two.  For rest of this paper, we shall use atomic units\footnote{Here, ``atomic unit'' is used in an \emph{abstract} sense: we do not assume (nor care) whether $m$, $e$, $\epsilon$ correspond to true values in vacuum, or some effective value in medium.} where $\hbar$, $m$, $e$ and $4 \pi \epsilon$ all have a value of one.  This implies that the units of energy are hartrees $E_{\mathrm{h}}$ and the units of length are Bohr radii $a$:
\begin{align*}
  E_{\mathrm{h}} &\equiv m \left(\frac{e^2}{4 \pi \epsilon \hbar}\right)^2 &
  a &\equiv \frac{4 \pi \epsilon \hbar^2}{m e^2}
\end{align*}
This leaves us with $(N, \omega)$ as the only necessary parameters to specify the quantum dot system, with $\omega$ in units of $E_{\mathrm{h}} / \hbar$.

The behavior of each electron in isolation is described by its kinetic energy $t(\bm{r})$ and the potential energy due to the trap $u(\bm{r})$, which combine to form the standard harmonic oscillator Hamiltonian $h(\bm{p}, \bm{r})$:
\begin{align*}
  t(\bm{p}) &\equiv \frac{1}{2} \bm{p}^2 &
  u(\bm{r}) &\equiv \frac{1}{2} \omega^2 \bm{r}^2 &
  h(\bm{p}, \bm{r}) &\equiv t(\bm{p}) + t(\bm{r})
\end{align*}
Here, $\bm{p}$ is the linear momentum of the electron and $\bm{r}$ is its position relative to the center of the trap.  The pairwise interaction between two electrons at positions $\bm{r}$ and $\bm{r}'$ is simply the Coulomb interaction:
\begin{align*}
  v(\bm{r}, \bm{r}') \equiv \frac{1}{\left|\bm{r} - \bm{r}'\right|}
\end{align*}
Then, the full many-body problem is described by the Hamiltonian
\begin{align*}
  \hat H \equiv \hat H^{[1]} + \hat H^{[2]}
\end{align*}
where $\hat{H}^{[1]}$ and $\hat{H}^{[2]}$ are the one- and two-body parts of the Hamiltonian,
\begin{align} \label{eq:onetwobodyhamiltonian}
\hat{H}^{[1]} &\equiv \sum_{j = 1}^N h(\hat{\bm p}_j, \hat{\bm r}_j) &
\hat{H}^{[2]} &\equiv \sum_{j_1 = 1}^N \sum_{j_2 = 1}^{j_1 - 1} v(\hat{\bm r}_{j_1}, \hat{\bm r}_{j_2})
\end{align}
The operators $\hat{\bm p}_j$ and $\hat{\bm r}_j$ respectively denote momentum and position of the $j$-th particle.\footnote{The order in which particles are labeled is unimportant; it is used only for bookkeeping.}

The first step towards a solution is to solve \emph{just} the noninteracting part $\hat{H}^{[1]}$.  In this case, the many-body Hamiltonian reduces to $N$ independent single-particle Hamiltonians $\hat{h}$, each of the form:
\begin{align*}
  \hat{h} = \frac{1}{2} \hat{\bm{\nabla}}^2 + \frac{1}{2} \omega^2 \hat{\bm{r}}^2
\end{align*}
This is a 2-dimensional harmonic oscillator problem for a single particle with known analytic solutions.  There are various ways to construct the set of eigenfunctions: we shall use the Fock-Darwin states, which conserve the angular momentum $\hat{L}_{\mathrm{z}}$. In polar coordinates, they are given by\citep[Appx.\ A]{lohne2010coupled}
\begin{gather*}
  \phi^{\text{FD}}_{n, m_\ell}(\rho, \varphi)
  \equiv \sqrt\omega
  \phi^{\text{FDr}}_{n, |m_\ell|}(\sqrt \omega \rho)
  \phi^{\text{Lz}}_{m_\ell}(\varphi)
\end{gather*}
where $\rho$ denotes the radial coordinate, $\varphi$ denotes the angular coordinate,
\begin{align*}
  \phi^{\text{FDr}}_{n, m}(r) &\equiv
  \sqrt{\frac{2 \cdot n!}{(n + m)!}}
  {\mathrm e}^{-r^2 / 2} r^m L_n^{(m)}(r^2) \\
  \phi^{\text{Lz}}_{m_\ell}(\varphi) &\equiv
  \frac{1}{\sqrt{2 \pi}} {\mathrm e}^{i m_\ell \varphi}
\end{align*}
and $L_n^{(\mu)}$ denotes the generalized Laguerre polynomial of degree $n$
with parameter $\mu$.  The states are labeled by two quantum numbers: the
principal quantum number $n \in \{0, 1, \ldots\}$ and the angular momentum
projection $m_\ell \in \mathbb Z$.

@@

\begin{figure}
\includegraphics[width=.48\textwidth]{fig-shell-structure}
\caption{Structure of the lowest 12 single-particle states in a noninteracting
  harmonic oscillator of frequency $\omega = 1$.  Each box represents a
  single-particle state, and the arrows denote the spin of the states.}
\label{fig:shell-structure}
\end{figure}
The eigenvalues of the single-particle Hamiltonian $\hat{h}$ are
given by
\begin{align*}
  \varepsilon_{n, m_\ell, m_{\text s}} \equiv (2 n + |m_\ell| + 1) \omega
\end{align*}
which satisfies
$\hat{h} \ket{\phi_{n, m_\ell, m_{\text s}}} = \varepsilon_{n,
  m_\ell, m_{\text s}} \ket{\phi_{n, m_\ell, m_{\text s}}}$.  Note that the
states are degenerate with respect to the spin projection $m_{\text s}$ since
our Hamiltonian $\hat{h}$ does not distinguish between them.  From
this equation, we can deduce the shell structure as shown in Figure
\ref{fig:shell-structure}.  It is notable that there are degeneracies beyond
that of the spin pairs: each shell contains a set of states that satisfy the
equation $2 n + |m_\ell| = k$ for some nonnegative integer $k$, which we
conveniently call the \textit{shell index}.  The shells are equidistant from
each other, with a spacing equal to the frequency $\omega$.

The many-body Hamiltonian $\hat{H}^{[1]}$ is the sum of the single-particle
Hamiltonians $\hat{h}_p$ over all particles $p$, where $p$ may now be
treated as a tuple of all three single-particle quantum numbers:
$(n, m_\ell, m_{\text s})$.  From the states $\phi$ in
\eqref{eq:singleparticlestate} we can construct the an eigenstate of the
many-body Hamiltonian $\hat{H}^{[1]}$ in the form of a Slater determinant
$\varnothing_{p_1, \ldots, p_N}$:
\begin{align*}
  \varnothing_{p_1, \ldots, p_N}(\bm r_1, \ldots, \bm r_N) \equiv
  \frac{1}{\sqrt{N}} \left|
  \begin{matrix}
    \phi_{p_1}(\bm r_1) & \cdots & \phi_{p_N}(\bm r_1) \\
    \vdots & \ddots & \vdots \\
    \phi_{p_1}(\bm r_N) & \cdots & \phi_{p_N}(\bm r_N) \\
  \end{matrix}
  \right|
\end{align*}
This automatically satisfies the Pauli exclusion principle, ensuring that the
Fermi statistics are satisfied.  The energy of these states is given by the
sum of the single-particle energies:
\begin{align*}
  E^{[1]}_{p_1, \ldots, p_N} = \varepsilon_{p_1} + \cdots + \varepsilon_{p_N},
\end{align*}
which satisfies $\hat{H}^{[1]} \ket{\varnothing_{p_1, \ldots, p_N}} = E^{[1]}_{p_1, \ldots, p_N} \ket{\varnothing_{p_1, \ldots, p_N}}$.  For ground states, we may have degeneracies due to the shell structure of the single-particle basis.  However, if $N = k_{\text F} (k_{\text F} + 1)$ for some nonnegative integer $k_{\text F}$, there are just enough particles to form a closed-shell system.  In this case, there is a unique, well-isolated ground state.  The values of $N$ for which this occurs are typically referred to as \textit{magic numbers} and we call $k_{\text F}$ the \textit{number of filled shells} (or ``Fermi level'').  This completes the solution of the noninteracting problem.

@@

Operators can be rewritten in this formalism as well.  The one-body
Hamiltonian $H^{[1]}$ defined by \eqref{eq:onetwobodyhamiltonian} can be
rewritten in second quantization as
\begin{align} \label{eq:second_quantized_one_body_hamiltonian}
  \hat H^{[1]} = \sum_{p, q} H_{p, q}^{[1]} \hat a_p^\dagger \hat a_q^{}
\end{align}
where the \textit{matrix element}
$H_{p, q}^{[1]} \equiv \bra{\varnothing_p} \hat{H}^{[1]} \ket{\varnothing_q}$.  This
can likewise be extended to $\hat{H}^{[2]}$,
\begin{align} \label{eq:second_quantized_two_body_hamiltonian}
  \hat{H}^{[2]} = \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s}^{}
  \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}
\end{align}
where $U_{p, q} \equiv \langle \varnothing_p | \hat U | \varnothing_q \rangle$ and
$V_{p, q, r, s} \equiv \langle \varnothing_{p, q} | \hat{H}^{[2]} | \varnothing_{r, s} \rangle$.
The complete Hamiltonian is given by their sum:
\begin{align}
  \hat{H} = \sum_{p, q} H^{[1]}_{p, q} \hat{a}_p^\dagger \hat{a}_q + \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_r \hat{a}_s  \label{eq:vacuumhamiltonian}
\end{align}
Note that due to the antisymmetry present in the Slater determinants, we have
the property that exchanging either $p$ with $q$ or $r$ with $s$ causes
$V_{p, q, r, s}$ to reverse in sign, leading to a $4$-fold symmetry that is
canceled out by the factor of $1/4$.  For this reason, the two-body matrix
element $V_{p, q, r, s}$ is commonly referred to as the \emph{antisymmetrized}
matrix element.  This is not to be confused with the non-antisymmetric
interaction integral defined by:
\begin{align*}
  \tilde V_{p, q, r, s} \equiv \int \phi_p(\bm r) \phi_q(\bm r') v(\bm r, \bm r') \phi_r(\bm r) \phi_s(\bm r') \D \bm r \D \bm r'
\end{align*}
The antisymmetrized matrix element is computed from the interaction integral
by $V_{p, q, r, s} = \tilde V_{p, q, r, s} - \tilde V_{p, q, s, r}$.  In our
work, we compute the interaction numerically via the OpenFCI software, which
numerical integrates the Coulomb interaction in the center-of-mass frame
[[cite Kvaal's paper; also, is the description correct?]].

@@

Since the coefficient matrix uniquely determines the unknown states, the variational problem has been reduced to that of solving of the coefficients $\bm C$ that minimize the energy $E_{\Phi}$, which is given by
\begin{align}
  E_{\Phi} &= \sum_{j} n_j H^{[1] \prime}_{j, j} + \sum_{j, k} n_j n_k V'_{j, k, j, k} \label{eq:hfenergy}
\end{align}
where the transformed operators are defined as
\begin{align}
  H^{[1] \prime}_{p, q} &\equiv \sum_{r, s} C_{r, p}^* H_{r, s}^{[1]} C_{s, q}^{} \label{eq:hftransform1} \\
  V'_{p, q, r, s} &\equiv \sum_{t, u, v, w} C_{t, p}^* C_{u, q}^* V_{t, u, v, w}^{} C_{v, r}^{} C_{w, s}^{} \label{eq:hftransform2}
\end{align}
and the occupancy number $n_p$ is defined as
\begin{align}
  n_p \equiv \begin{cases}
    1 & \text{if $p$ is an occupied state in $\Phi$} \\
    0 & \text{if $p$ is an unoccupied state in $\Phi$}
  \end{cases}
\end{align}
and further define $\bar n_p \equiv 1 - n_p$ as the complement.

With the method of Lagrange multipliers, the minimization problem can be reduced to the solving of a nonlinear equation -- the \textit{Hartree-Fock equation}:
\begin{align} \label{eq:hartreefock}
  \bm F \bm C = \bm C \bm \epsilon
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p, q} \equiv H_{p, q}^{[1]} + \sum_{r, s, j} n_j C_{r, j}^* V_{p, r, q, s}^{} C_{s, j}^{}
\end{align}
and $\bm \epsilon$ is a vector of Lagrange multipliers with each element $\epsilon_q$ associated with a particular state $\phi'_q$.

@@

To make use of the HF solution as the reference state for future calculations, we transform the operators via \eqref{eq:hftransform1} and \eqref{eq:hftransform2}.  Therefore, the post-HF methods described in the next few sections generally use the transformed operators $\hat{H}^{[1] \prime}$ and $\hat{H}^{[2] \prime}$ as inputs.  However, for generality, we omit the prime symbols as the methods do not necessitate a state optimized by HF.

@@

We may then define \emph{normal ordering with respect to the reference state} $\Phi$, a notational device that rearranges a string of operators such that every quasiparticle creation operator appears before a quasiparticle annihilation operator by repeatedly exchanging neighboring operators, with each exchange incurring a change in sign.  A simple example would be the normal ordering of a pair of field operators such as:
\begin{align*}
  \normord{\hat{b}_p \hat{b}_q^\dagger} =
  -\hat{b}_q^\dagger \hat{b}_p
\end{align*}
Here, we denote normal ordering of an operator string with respect to $\Phi$ by a pair of colons.  The expectation value of a normal ordered string of operators with respect to $\Phi$ is guaranteed to vanish:
\begin{align*}
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q \ket{\Phi} = 0 &
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q^\dagger \hat{b}_s \hat{b}_r \ket{\Phi} = 0
\end{align*}
The normal ordering of the quasiparticle field operators are straightforward to compute, but what is more interesting is the normal ordering of the physical field operators $\hat a$ and $\hat a^\dagger$ with respect to $\Phi$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q} &=
  \hat{a}_p^\dagger \hat{a}_q - n_p \delta_{p, q}
\end{align*}
The normal ordering notation thus provides a concise way of expressing the right-hand side.  In this new notation, we can rewrite the Hamiltonian $\hat H = \hat{H}^{[1]} + \hat{H}^{[2]}$ in its \emph{normal-ordered representation}:
\begin{align}
  \hat{H} = H^{[0]} + \sum_{p, q} H^{[1]}_{p, q} \normord{\hat{a}_p^\dagger \hat{a}_q} + \frac{1}{4} \sum_{p, q, r, s} H^{[2]}_{p, q, r, s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_r \hat{a}_s}
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  H^{[0]} &\equiv \sum_{i} n_i H^{[1]}_{i, i} + \frac{1}{2} \sum_{i, j} n_i n_j V_{i, j, i, j} \\
  H^{[1]}_{p, q} &\equiv H^{[1]}_{p, q} + \sum_{i} n_i V_{p, i, q, i} \\
  H^{[2]}_{p, q, r, s} &\equiv V_{p, q, r, s} \\
\end{align*}
If $\Phi$ is a state optimized by the HF method, then $H^{[0]}$ is simply the HF energy $E_\Phi$ and $H^{[1]}_{p, q}$ is the Fock matrix $F_{p, q}$.  The normal-ordered representation of $\hat H$ in \eqref{eq:normordhamiltonian} is completely equivalent to that of $\hat H$ in \eqref{eq:vacuumhamiltonian}.  The only difference is that it rearranges the matrix elements so as to emphasize the reference state $\Phi$ rather than vacuum state $\ket{\varnothing}$.  This makes a critical difference when the expressions containing 3-body or higher terms are \emph{truncated} to only 2-body.  If the reference state $\Phi$ is good approximation of the true solution $\Psi$, then the importance of the higher-body parts of the Hamiltonian are greatly reduced.

@@

In QDPT, the solutions of the approximate Hamiltonian $\hat{H}^{[1]}$ are used as the basis of the model space.  One then assumes the existence of an operator $\hat \Omega$, known as the \textit{wave operator}, that maps some set of states $\tilde \Psi^{\mathrm{o}}_u$ within the model space to the exact ground state $\Psi_u$:
\begin{align} \label{eq:omega-condition1}
  \Psi_u = \hat \Omega \tilde \Psi^{\mathrm{o}}_u
\end{align}
The states $\tilde \Psi^{\mathrm{o}}_u$ consist of some mixture of the eigenstates
$\Psi^{\mathrm{o}}_{u'}$ of the approximate Hamiltonian $\hat{H}^{[1]}$.

There is some freedom in the choice of the wave operator $\hat \Omega$.  We
assume it has the following form:
\begin{align} \label{eq:omega-condition2}
  \hat \Omega = \hat P + \hat Q \hat \Omega \hat P
\end{align}
where $\hat P$ projects any state into the model space and $\hat Q$ is the
complement of $\hat P$.  This entails that the exact states $\Psi_u$ is no
longer normalized but instead satisfies the typical intermediate
normalization: $\langle \Psi_u | \tilde \Psi^{\mathrm{o}}_u \rangle = 1$.
